Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters

### 2019-05-19 paper.tex
added figures and text. need to complete discussion

    \tristan{add a paragraph to explain master, worker, driver}

    Spark has many configuration options available for dynamic resource provisioning.
    Its standalone cluster provides two deployment modes: client and cluster. The client
    deploy mode executes the driver within the \texttt{spark-submit} process as
    a client to the cluster. The cluster deploy-mode, however, runs the driver within
    a worker process. Such a configuration is practical when the driver program
    is deployed on a machine not co-located in the cluster to reduce network 
    latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
    feature which provides fault-tolerance to the driver \tristan{, by means of checkpointing?}.
    This is particularly useful for
    pilot-based overlay clusters where the running driver may fail due to walltime expiration.
    Cluster deploy mode is not available for Python applications using the standalone mode \tristan{standalone mode wasn't defined.
    Rewrite positively saying that it's ony available in Scala?}.

@@

    Spark also provides master fault tolerance through single-node recovery, where a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder. Although we start a master on each pilot, we have not incorporated this recovery function
    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
    were any nodes can fail by design. \tristan{Last two sentences should be moved to job templates.}

    %%maybe add checkpointing

    \subsection{Job Templates}

    \tristan{This section could be renamed to ``SPA system design'', and you could add a paragraph explaining
    that for the sake of the experiment, you reimplemented the simplest possible pilot system since existing ones (refs)
    were either not found online, or difficult to use.}

    Two different job templates were used to implement the two main
    conditions compared in our experiments: the batch submission
    template and the pilot submission template. The batch submission template was
    inspired by the template provided by Compute Canada to launch Spark applications 
    on Slurm, \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
    The template operates as follows: certain resource requirements are requested by
    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
    of nodes and number of tasks per node). Once these resources are allocated, a 
    master is started on one of the requested node resources. Then, after the master
    has successfully started, the workers are started on all nodes. Multiple worker
    instances are started on a single node by setting the \texttt{SPARK\_WORKER\_INSTANCES}
    environment variable to the number of tasks per node. Each worker is given as many
     cores as specified by the user in the Slurm resource allocation request.
    After both the masters and the workers have successfully started, the driver is finally
    started. The amount of memory given to each executor corresponds to 95\% of the 
    available memory on the node, to allow for offheap space. 
    The Spark deploy-mode selected for the batch template is client-mode \TG{why?}.

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. Only one pilot,
    the first pilot to attempt to launch the driver, will do so \tristan{will do what? launch the driver?}.
    The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}
    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
    description clearer.}

@@

    \todo{walltime parameter of workflow}

@@

    \todo{perhaps expand a bit here \TG{I have a formal proof for eq (1) but I don't think
    it's required here.}}

    \subsection{Added value of pilot scheduling}
        \begin{table}                                                                    
            \centering                                                                       
            \begin{tabular}{c|c|c}                                                             
            \rowcolor{headcolor}                                                             
            Configuration & RAM (GB) & Tasks \TG{Should be cores?}\\                               
            \hline                                                                           
            1 & 112 & 16 \\                                               
            2 & 224 & 32 \\                                               
            3 & 336 & 48 \\
            4 & 448 & 64 \\
            \end{tabular}                                                                    
            \setlength{\belowcaptionskip}{-10pt}                                             
            \caption{Resource configurations}                                                    
            \label{table:dedicatednodes}                                                            
        \end{table}

@@

       As the focus of our experiments here is to measure the impact pilot scheduling
       has on queueing times, we wanted to ensure that our walltime estimates would
       be consistent for each experimental condition. Therefore, we adjusted task 
       delay based on the maximum level of parallelism to ensure that walltimes would
       not need to be readjusted for each configuration. Given utmost parallelism,
       the sleep delay added would amount to a makespan of 1 hour regardless
       of configuration. The actual processing time of the application, however, 
       would differ between configurations as we did not account for task duration (incrementation time, line
       8 in Algorithm~\ref{alg:incrementation})
       in the sleep delay. However, task duration was not expected to significantly impact
       overall processing time, but nevertheless, we set application walltime to 
       2 hours and 30 minutes for all configurations to account for additional
       processing time that may be incurred in applications with lower levels of 
       parallelism. \TG{this last sentence is a bit confusing}

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on B\'eluga and Cedar. In B\'eluga (Figure~\ref{fig:beluga1}) \tristan{why just in 1a and not 1b, ...?}, 16 pilots were always 
slower than batch except for repetitions 4 and 12. The maximum speedup achieved
was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
pilots for the majority of the cases, except a couple, in which the speedup brought
by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
that we see a maximum speedup of 2.8, however, it generally remains that
pilots are slightly slower. \TG{You should report the average speed-up of
pilots: 8 pilots vs batch and 16 pilots vs batch.}

@@

The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
the makespan of batch scheduled applications are generally faster than that of pilots,
with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
varies between the iterations. In general, both pilot configurations performed the
same.

@@

In order to determine whether queueing time was the cause of the difference 
in makespans between batch and pilots \tristan{rather than other overheads?}, we compared makespan with queuing time of 
first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
makespans is primiarily due to queueing time. \tristan{I don't like these figures: if you want to show
this, you should plot the makespan vs the queuing time of the first pilot, in an XY plot. But then the MW plot
is better because it takes into account the queuing time of all pilots, not only the first one. I would remove this figure.}

      \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
            \caption[]%
            {{\small Makespan}}
            \label{fig:makespancedar}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/pq_4_cedar}
            \caption[]%
            {{\small Queuing time}}
            \label{fig:queuecedar}
        \end{subfigure}
        \caption[]
        {\small Side-by-side comparison of total makespan and queueuing time of
        first pilots or batch application for Configuration 4 running on Cedar}
        \label{fig:mqcedar}
    \end{figure*}

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our performance model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application, in seconds
    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
    \item $d$ is the task delay associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
\tristan{You should remove the ceil from the plot too.}
As can be seen in the figure, although pilots and batch had similar makespans
for a given number of average workers, pilots were consistently slower than batch
for the same number of workers. Both batch and pilots were consistent with the 
model, however \TG{there is a contradiction here: if both batch and pilots were consistent with the model, then
they should have similar makespans.}. In general, both pilots and batch had the same number of average
workers. \TG{You could also plot the difference in the number of workers, for each experiment.}

@@

%Technical overhead of using pilots + an overlay cluster over a cluster.
Pilots do not appear to provide any performance advantage over batch scheduling.
This can be due to a few reasons. The queuing time for batch and pilots were similar, 
which can be an indication that the required resources to run the batch requests were
available, and our priority on the cluster was what affected the queuing time \TG{I don't understand this, could you clarify?}. This may, in turn,
explain why batch was sometimes significantly slower than pilots -- queueing time was not
only controlled by priority, but also by available resources.

>>>>

    There are three possible cluster managers available to use in Spark applications:
    the Standalone cluster manager, YARN and Mesos. The Standalone cluster manager is the
    most basic cluster manager available and is packaged directly with Spark. YARN is a
    resource manager designed for the resource management of Hadoop applications. However,
    it can be used with different types of applications as well. In contrast, Mesos was 
    designed to manage the resources for a variety of applications. As such, it incorporates
    strategies that are more efficient for use with different frameworks making it possible
    to be used as an HPC scheduler. Due to Standalone's lightweight nature and our intent
    to start Spark clusters per application, we focus on Standalone mode.

    A Spark cluster is made up of three components: the cluster manager, workers and
    driver. The cluster manager, also known as the Spark master in Standalone mode, is responsible for the
    resource provisioning within the cluster. Unlike the workers and driver, it is not
    application specific. The driver, which can start in the client where it is launched (i.e. client deploy mode)
    or directly on a worker node (i.e. cluster deploy mode), must request resources from the master
    to launch its tasks. The master the requests workers with the necessary resources available to start
    executors. The driver can then communicate with the executors to schedule and launch its tasks. Both the driver and
    the executors are application specific and connected directly to the master.

    Spark has many configuration options available for dynamic resource provisioning.
    Its standalone cluster provides two deployment modes: client and cluster. The client
    deploy mode executes the driver within the \texttt{spark-submit} process as
    a client to the cluster. The cluster deploy-mode, however, runs the driver within
    a worker process. Such a configuration is practical when the driver program
    is deployed on a machine not co-located in the cluster to reduce network 
    latency between workers and driver. The cluster deploy mode also has the \texttt{supervise}
    feature which allows the driver to be restarted in case of failure.
    This is particularly useful for
    pilot-based overlay clusters where the running driver may fail due to walltime expiration.
    Using the Standalone cluster manager, client mode is supported in all APIs, whereas cluster mode is
    only available for applications using the Scala, Java and R APIs.

@@

    Spark also provides master fault tolerance through single-node recovery, where a shared
    filesystem is available. To enable this, a recovery folder must be set. Should the master
    fail, a new master will be able to takeover based on the information available in the recovery
    folder.

    In long running applications, it is necessary to be able to recovery the application execution from
    last successful state. Spark provides a checkpointing mechanism in order to ensure that the application
    is fault-tolerant to any kind of non application-related failure, such as a system failure.

    \subsection{SPA system design}

    There is not shortage of scripts available online for starting a Spark cluster on HPC~\footnote{\url{https://github.com/NIH-HPC/spark-slurm},}\footnote{\url{https://www.sherlock.stanford.edu/docs/software/using/spark/},}\footnote{\url{https://sparkhpc.readthedocs.io},}\footnote{\url{https://www.osc.edu/~troy/pbstools/man/pbs-spark-submit}}.
    However, all these scripts start up a cluster within a single batch call. At most, there are two calls, one to startup the cluster,
    and the second, to start the application execution. None of these use any kind of pilot-scheduling approach, nor do they discuss the
    effects of queuing time and how they might want to adjust it in the case where the driver is started separately
    from the batch job to start the cluster. Even in the case of RADICAL-Pilot, as per some available tutorials for using the pilot-job application with
    Spark~\footnote{\url{https://github.com/radical-cybertools/pilot-streaming/blob/master/examples/Pilot-Streaming-GettingStarted.ipynb},}\footnote{\url{https://github.com/radical-cybertools/MIDAS-tutorial/blob/master/pilot/Pilot-Spark.ipynb}}, it appears that the Spark cluster is started within a single pilot and the application is submitted, in a separate process, to that pilot.
    Due to the fact that existing solutions mainly rely on a single batch call to startup the Spark
    cluster and do not appear to be able to submit multiple smaller jobs to improve queuing times, we have decided to implement our own solution.

    Two different job templates were used to implement the two main
    conditions compared in our experiments: the batch submission
    template and the pilot submission template. The batch submission template was
    inspired by the template provided by Compute Canada to launch Spark applications 
    on Slurm, \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
    The template operates as follows: certain resource requirements are requested by
    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
    of nodes and number of tasks per node). Once these resources are allocated, a 
    master is started on one of the requested node resources. Then, after the master
    has successfully started, the workers are started on all nodes. Multiple worker
    instances are started on a single node by setting the \texttt{SPARK\_WORKER\_INSTANCES}
    environment variable to the number of tasks per node. Each worker is given as many
     cores as specified by the user in the Slurm resource allocation request.
    After both the masters and the workers have successfully started, the driver is finally
    started. The amount of memory given to each executor corresponds to 95\% of the 
    available memory on the node, to allow for offheap space. 
    The Spark deploy mode selected for the batch template is client mode. We selected this
    deploy mode over cluster mode as driver recovery would not be required in batch in the
    case of walltime expiration, as all the allocated resources will expire at the same time.
    Moreover, many scientific applications are written in Python, which cannot use cluster mode using
    the Standalone resource manager. Should pilot scheduling using cluster mode significantly reduce
    queuing time compared to batch scheduling using client mode, it would provide enough justification to extend
    Spark's Standalone scheduler to include cluster mode for PySpark applications.

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. However, there will only be one
    pilot which will start the driver process. The pilot selected to start the driver is the first one
    that attempts to do so by way of lockfile.
    The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \texttt{supervise} mode, which is only available in cluster deploy. As
    with the masters, it is particularly important to have a fault-tolerant
    driver in pilot-scheduling scenarios due to possible walltime
    expiration.
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.
    % \TG{It should be explained before that one of the interests
    % of pilots is to ``play'' with shorter walltimes}. \TG{It would also be
    % nice to have runs where pilots 
    % have shorter walltimes, to see if queuing times are shorter too.}
    \TG{You could add a figure showing the driver, workers, master and stand-by masters, to make the system 
    description clearer.}

    Although we start a master on each pilot, we have not incorporated the master recovery function
    for our experiments. This feature is, however, important and should be incorporated in pilot schedulers
    were any nodes can fail by design.

@@

@@

    \subsection{Added value of pilot scheduling}
        \begin{table}                                                                    
            \centering                                                                       
            \begin{tabular}{c|c|c|c}                                                             
            \rowcolor{headcolor}                                                             
            Configuration & RAM (GB) & Tasks & Cores per task\\                               
            \hline                                                                           
            1 & 112 & 16 & 1\\                                               
            2 & 224 & 32 & 1\\                                               
            3 & 336 & 48 & 1\\
            4 & 448 & 64 & 1\\
            \end{tabular}                                                                    
            \setlength{\belowcaptionskip}{-10pt}                                             
            \caption{Resource configurations}                                                    
            \label{table:dedicatednodes}                                                            
        \end{table}

@@

       As the focus of our experiments here is to measure the impact pilot scheduling
       has on queueing times, we wanted to ensure that our walltime estimates would
       be consistent for each experimental condition. Therefore, we adjusted task 
       delay based on the maximum level of parallelism to ensure that walltimes would
       not need to be readjusted for each configuration. Given utmost parallelism,
       the sleep delay added would amount to a makespan of 1 hour regardless
       of configuration. The actual processing time of the application, however, 
       would differ between configurations as we did not account for task duration (incrementation time, line
       8 in Algorithm~\ref{alg:incrementation})
       in the sleep delay. Due to slight variability in task durations between experiments, in 
       addition to variability in application parallelism due to queuing times of each pilot Slurm job, 
       we have set the walltime to be 2h30 to ensure experiment completion.

@@

As can be seen in both Figures~\ref{fig:makespansbeluga} and~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on either system. In B\'eluga and Cedar, 16 pilots were generally slower than batch except for a 
few occurrencs. When 16 pilots were faster, it was generally either very slight, or, as expected,
when batch queuing times were very large. The maximum speedup achieved
was approximately 5.6~x for 8 and 16 pilots on B\'eluga and 2.6~x on Cedar. 
It is only in Configuration~3 (Figure~\ref{fig:beluga4})
that we spot any large speedups, however, it generally remains that
pilots are slightly slower, even in Configuration~3. The same can be said about Cedar, however,
it occured with Configuration~4 instead of 3. On average, pilots were up to 0.11~x slower on B\'eluga and
0.17~x slower on Cedar (see Table~\ref{table:speedup}), both of which occurred when using 16 pilots.
The only times when pilots were found to be, on average, faster, was in Configuration~3 on B\'eluga
and Configuration~4 on Cedar where the speedup was found to reach a max of 1.46 for 8 pilots on 
B\'eluga.

@@

@@

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our performance model. Makespan for the model was calculated as:
$$
M = \frac{125\times(d + 20)}{W}\times 10,
$$
where:
\begin{itemize}
    \item $125$ is the number of chunks BigBrain was split into
    \item $M$ is the makespan of the application, in seconds
    \item $W$ is the average number of Spark workers throughout the execution, computed as in Equation~\ref{eq:avgw}
    \item $d$ is the task delay associated with a given configuration
    \item $20$ is the average measured incrementation duration, in seconds, for a BigBrain chunk
    \item $10$ is the number of iterations
\end{itemize}
As can be seen in the Figure~\ref{fig:mwall}, although pilots and batch had similar makespans
and followed the general trend denoted by the model, pilots were consistently slower than batch
for the same number of average workers. When comparing the average number of worker difference
(Figures\ref{fig:nworkersbeluga} and \ref{fig:nworkerscedar}), it can be seen that pilot average
number of workers start to exceed batch and the requirements increase. However, employing 16 pilots
is generally always more costly than 8 pilots.

    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_beluga}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwbeluga1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_beluga}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwbeluga2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_beluga}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwbeluga3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_beluga}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwbeluga4}
        \end{subfigure}
        \caption[]
        {\small The average worker difference on B\'eluga between batch and pilots for all configurations and repetitions}
        \label{fig:nworkersbeluga}
    \end{figure*}
    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_1_cedar}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:nwcedar1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_2_cedar}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:nwcedar2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_3_cedar}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:nwcedar3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/nworkers_4_cedar}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:nwcedar4}
        \end{subfigure}
        \caption[]
        {\small The average worker difference on Cedar between batch and pilots for all configurations and repetitions}
        \label{fig:makespanscedar}
    \end{figure*}

\begin{table}                                                                    
    \centering                                                                       
    \begin{tabular}{c|c|c|c|c}                                                             
        {} & \multicolumn{2}{c}{B\'eluga} & \multicolumn{2}{c}{Cedar}\\
    \rowcolor{headcolor}                                                             
    Configuration & 8 pilots & 16 pilots & 8 pilots & 16 pilots\\                               
    \hline                                                                           
    1 & 0.949 & 0.932 & 0.724 & 0.874\\                                               
    2 & 0.988 & 0.916 & 0.836 & 0.826\\                                               
    3 & 1.458 & 1.369 & 0.941 & 0.964\\
    4 & 0.970 & 0.891 & 1.049 & 1.068\\
    \end{tabular}                                                                    
    \setlength{\belowcaptionskip}{-10pt}                                             
    \caption{Speedup of pilots}                                                    
    \label{table:speedup}                                                            
\end{table}

@@

%Technical overhead of using pilots + an overlay cluster over a cluster.
Pilots do not appear to provide any performance advantage over batch scheduling.
This can be due to a few reasons. The queuing time for batch and pilots were similar, 
which can be an indication that we were just lacking the right level of priority to 
have been scheduled immediately. In the instances where batch was found to be significantly slower,
it might have been then case that we had obtained the necessary priority level but there
were not enough resources available to be scheduled.
